Panel Seminars

### 2022-06-24 DGE_comparison_v2.Rmd
Rendered site to include RNAseq in navigation, made minor edits to RADseq, changed to DEG_comparision_v2

---
title: "Fitting Multifactorial Models of Differential Expression"
author: "Samuel N Bogan"
bibliography: DGE_Comp.bib
---

#Intro to multifactorial RNA-seq models

@@

##Targeted audience
|           This walkthrough is for both intermediate and expert practitioners of RNA-seq analyses, and most of all, those new to multifactorial RNA-seq. We have written this page with the assumption that readers have a basic familiarity with model fitting and statistical analysis, as well as next generation sequencing technologies, RNA-seq workflows, and the R statistical environment.

##Variables we will cover

@@

##Packages we will cover

@@

#Features of popular DGE packages

@@

##Summary of approach

@@

```{r setup, include = FALSE}

# Set root directory
knitr::opts_knit$set( root.dir = '~/Documents/GitHub/DGE_comparison/' )

@@

#invisible(lapply(c("DESeq2", "edgeR", "tidyverse", "ape", "vegan", "GGally",
                   #"arrayQualityMetrics", "rgl", "adegenet", "MASS",
                   #"data.table", "plyr", "lmtest", "reshape2", "Rmisc", "lmerTest"),
                 #function(p){
                   #if(! p %in% rownames(installed.packages())) {
                     #install.packages(p)
                  # }
                  # library(p, character.only=TRUE)
                 #}))

@@

#Filter and visualize read counts

@@

# Read in matrix of RSEM expected read counts
data <- read.delim("Input_data/GE2_expected_counts_gene.matrix", sep = "\t", header = T, row.names = 1)

@@

##MDS plot visualizing experimental factors

@@

#Non-linear effects

@@

##Non-linear effects: example in edgeR

@@

##Plotting non-linear effects

@@

#Interactive effects

@@

##Interactive effects: edgeR

@@

##Interactive effects: limma-Voom

@@

##Interactive effects: DESeq2

@@

#Comparing test statistics: interactive effects

@@

#Random Intercepts: example in Voom

@@

#Random slopes

@@

#Conclusion: Moving forward with multifactorial RNA-seq

@@

#References

>>>>

```{r,message = FALSE,echo = FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE)
opts_chunk$set(fig.width = 10,
               fig.height = 5,
               cache = FALSE)
```
---
title: "Fitting Multifactorial Models of Differential Expression"
author: "Samuel N Bogan"
bibliography: DGE_Comp.bib
---

# Intro to multifactorial RNA-seq models

@@

## Targeted audience  
|           This walkthrough is for both intermediate and expert practitioners of RNA-seq analyses, and most of all, those new to multifactorial RNA-seq. We have written this page with the assumption that readers have a basic familiarity with model fitting and statistical analysis, as well as next generation sequencing technologies, RNA-seq workflows, and the R statistical environment.

## Variables we will cover

@@

## Packages we will cover

@@

# Features of popular DGE packages

@@

## Summary of approach

@@

```{r setup, include = FALSE}
# should remove from Rmd
# Set root directory
#knitr::opts_knit$set( root.dir = '~/Documents/GitHub/DGE_comparison/' )

@@

#invisible(lapply(c( "tidyverse", "ape", "vegan", "GGally",
                   #, "rgl", "adegenet", "MASS",
                   #"data.table", "plyr", "lmtest", "reshape2", "Rmisc", "lmerTest","statmod"),
                 #function(p){
                   #if(! p %in% rownames(installed.packages())) {
                     #install.packages(p)
                  # }
                  # library(p, character.only=TRUE)
                 #}))

#if (!require("BiocManager", quietly = TRUE))
    #install.packages("BiocManager")

#BiocManager::install(c("DESeq2","edgeR","arrayQualityMetrics"))

@@

# Filter and visualize read counts

@@

# Read in matrix of RSEM expected read counts
data <- read.delim("RNA_data/GE2_expected_counts_gene.matrix", sep = "\t", header = T, row.names = 1)

@@

## MDS plot visualizing experimental factors

@@

# Non-linear effects

@@

## Non-linear effects: example in edgeR

@@

## Plotting non-linear effects

@@

# Interactive effects

@@

## Interactive effects: edgeR

@@

## Interactive effects: limma-Voom

@@

## Interactive effects: DESeq2

@@

# Comparing test statistics: interactive effects

@@

# Random Intercepts: example in Voom

@@

# Random slopes

@@

# Conclusion: Moving forward with multifactorial RNA-seq

@@

# References

### 2022-06-24 RADseq.Rmd
Rendered site to include RNAseq in navigation, made minor edits to RADseq, changed to DEG_comparision_v2

@@

|  | [dDocent](https://www.ddocent.com/) | [ipyrad](https://ipyrad.readthedocs.io/en/latest/index.html)| [Stacks2](https://catchenlab.life.illinois.edu/stacks/) |
|---|---|---|---|
| Supported datatypes | Paired-end: ddRAD, ezRAD, RAD (random shearing), data with large overlap  between forward and reverse reads  
Single-end: any RADseq method  de novo and reference-based.  
If doing de novo assembly, reads cannot be trimmed outside of dDocent |  |  |
| Unique aspects | Novel data reduction approach used to inform coverage cutoffs |  |  |
| Documentation | Very good (esp. tutorials), active community support on Google Groups | Excellent (esp. installation, parameter explanations, and tutorials), active community support on Gitter | Very good, active community support on Google Groups |
| Speed/Accuracy | fastest and most accurate | close to dDocent accuracy, more over-splitting is possible  if parameters are not tuned | produces some untrue genome fragments (esp. with higher levels of indel polymorphism), but can be addressed with downstream filtering |
| Open source/development | Open source | Open source | no |
| Filtering options | Minimal default filtering as implemented in VCFtools. Ideal for those who want full freedom in filtering their SNPs. |  |  |
| Output options | Only produces a VCF, fasta of the de novo assembly, and individual BAM files  from mapping to the reference  |  |  |
| Popgen analyses | No additional popgen analyses included  |  |  |
|  |  |  |  |

@@

To plot the missingness on the command line, you can use gnuplot:
```{bash}
#code from Jon Puritz
mawk '!/IN/' RAD_data/OL_subset.imiss | cut -f5 > totalmissing
gnuplot << \EOF
set terminal dumb size 120, 30
set autoscale
unset label
set title "Histogram of % missing data per individual"
set ylabel "Number of Occurrences"
set xlabel "% of missing data"
#set yr [0:100000]
binwidth=0.01
bin(x,width)=width*floor(x/width) + binwidth/2.0
plot 'totalmissing' using (bin($1,binwidth)):(1.0) smooth freq with boxes
pause -1
EOF
```

@@

There are many parameter choices to make when assembling and genotyping, and those parameters vary slightly across pipelines. The parameters below are general across RADseq pipelines (although with different names), and are shown to have the largest impact on genoyping error rates. 
### Clustering threshold {#cluster}
#### (de novo assembly only)
During *de novo* RADseq assembly, reads are assembled into contigs (contiguous sequences), so that each contig represents a single locus. This assembly can occur separately for each individual (ipyrad) or jointly for multiple individuals (dDocent, Stacks). To call SNPs, in the case of ipyrad, contigs are then clustered and aligned across individuals, while in dDocent and Stacks, reads for each individual are mapped to the "reference" contigs. All of these pipelines use sequence similarity (or **clustering threshold**) to determine if fragments are homologous (from the same place in the genome) and therefore should be clustered together.

@@

`r colorize("This section is still under development! Please contribute on Github or provide feedback in [Discussions](https://github.com/MarineOmics/marineomics.github.io/discussions)", "red")`
Even with a reduced representation sequencing approach, your assembly and genotyping pipeline will generate data for thousands of SNPs across your individuals. These SNPs will always need to be filtered before doing downstream analyses.  
#### Main reasons to filter your data:

>>>>

# Setup for running code

@@

|  | [dDocent](https://www.ddocent.com/) | [ipyrad](https://ipyrad.readthedocs.io/en/latest/index.html)| [Stacks2](https://catchenlab.life.illinois.edu/stacks/) |
|---|---|---|---|
| Supported datatypes | Paired-end: ddRAD, ezRAD, RAD (random shearing), data with large overlap between forward and reverse reads; Single-end: any RADseq method  de novo and reference-based. If doing de novo assembly, reads cannot be trimmed outside of dDocent | Wide range of RAD/GBS methods with paired/single end data, see [ipyrad documentation](https://ipyrad.readthedocs.io/en/master/4-data.html) | Paired-end and single-end data for single and double digest RAD and DART; GBS with single-end sequencing. NOT suitable for paired-end GBS. |
| Unique aspects | Novel data reduction approach used to inform coverage cutoffs | Python API for popgene and phylogenetic analyses written specifically for RAD data |  |
| Documentation | Very good (esp. tutorials), active community support on Google Groups | Excellent (esp. installation, parameter explanations, and tutorials), active community support on Gitter | Very good, active community support on Google Groups |
| Speed/Accuracy | fastest and most accurate | close to dDocent accuracy, more over-splitting is possible  if parameters are not tuned | produces some untrue genome fragments (esp. with higher levels of indel polymorphism), but can be addressed with downstream filtering |
| Open source/development | Open source | Open source, primarily combines existing software | No |
| Filtering options | Minimal default filtering as implemented in VCFtools. Ideal for those who want full freedom in filtering their SNPs. | Wide range of filtering options  | Moderate range of filtering options |
| Output options | Only produces a VCF, fasta of the de novo assembly, and individual BAM files  from mapping to the reference  | Lots of output formats (see [documentation](https://ipyrad.readthedocs.io/en/master/output_formats.html)) | Lots of Stacks-specific output and log files, as well as various inputs for popgen programs (see [documentation](https://catchenlab.life.illinois.edu/stacks/comp/populations.php)) |
| Popgen analyses | No additional popgen analyses included  | Unique, flexible Python API implementing range of popgen and phylogenetic methods (see [documentation](https://ipyrad.readthedocs.io/en/master/API-analysis/index.html))   | Popgen summary statistics (F-statistics, pi), including kernel-smoothing along reference genome. Underlying equations and assumptions for these are not very clear. |
|  |  |  |  |

@@

To plot the missingness on the command line, you can use gnuplot:
```{bash, eval = T}
#code from Jon Puritz
mawk '!/IN/' RAD_data/OL_subset.imiss | cut -f5 > totalmissing
gnuplot << \EOF
set terminal dumb size 120, 30
set autoscale
unset label
set title "Histogram of % missing data per individual"
set ylabel "Number of Occurrences"
set xlabel "% of missing data"
#set yr [0:100000]
binwidth=0.01
bin(x,width)=width*floor(x/width) + binwidth/2.0
plot 'totalmissing' using (bin($1,binwidth)):(1.0) smooth freq with boxes
pause -1
EOF
```

@@

There are many parameter choices to make when assembling and genotyping, and those parameters vary slightly across pipelines. The parameters below are general across RADseq pipelines (although with different names), and are shown to have the largest impact on genoyping error rates.

### Clustering threshold {#cluster}  
#### (de novo assembly only)  
During *de novo* RADseq assembly, reads are assembled into contigs (contiguous sequences), so that each contig represents a single locus. This assembly can occur separately for each individual (ipyrad) or jointly for multiple individuals (dDocent, Stacks). To call SNPs, in the case of ipyrad, contigs are then clustered and aligned across individuals, while in dDocent and Stacks, reads for each individual are mapped to the "reference" contigs. All of these pipelines use sequence similarity (or **clustering threshold**) to determine if fragments are homologous (from the same place in the genome) and therefore should be clustered together.

@@

`r colorize("This section is still under development! Please contribute on Github or provide feedback in [Discussions](https://github.com/MarineOmics/marineomics.github.io/discussions)", "red")`

Even with a reduced representation sequencing approach, your assembly and genotyping pipeline will generate data for thousands of SNPs across your individuals. These SNPs will always need to be filtered before doing downstream analyses.

#### Main reasons to filter your data:

### 2022-06-24 choosing-population-genetics.Rmd
Rendered site to include RNAseq in navigation, made minor edits to RADseq, changed to DEG_comparision_v2

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE)
library(knitcitations)
library(kableExtra)
opts_chunk$set(fig.width = 10,
               fig.height = 5,
               cache = FALSE)
cite_options(citation_format = "pandoc", max.names = 3, style = "html",
             hyperlink = "to.doc")
```

>>>>

```{r setup, include=FALSE}
#library(knitr)
knitr::opts_chunk$set(echo = TRUE)
library(knitcitations)
library(kableExtra)
knitr::opts_chunk$set(fig.width = 10,
               fig.height = 5,
               cache = FALSE)
cite_options(citation_format = "pandoc", max.names = 3, style = "html",
             hyperlink = "to.doc")
```

### 2022-06-24 index.Rmd
Push index.rmd and knit with multifac RNAseq page
Push knitted index.html after formatting DGE_comparison.html
Rendered site to include RNAseq in navigation, made minor edits to RADseq, changed to DEG_comparision_v2

* Transcriptomics with RNA-Seq (coming soon!)  
* Differential gene expression (coming soon!)

@@

#### [Katherine Silliman, PhD](https://ksilliman.weebly.com/), Auburn University - kes0132[at]auburn.edu {.unlisted .unnumbered}
#### [Sam Bogan](http://www.hofmannlab.com/sam-bogan.html), UC-Santa Barbara -snbogan[at]ucsb.edu {.unlisted .unnumbered}
#### [Alan Downey-Wall](https://cos.northeastern.edu/people/alan-downey-wall/), Northeastern University {.unlisted .unnumbered}
#### Chris Mantegna, University of Washington (Intern)  {.unlisted .unnumbered}
#### [Danielle Davenport](https://www.researchgate.net/profile/Danielle-Davenport-2), University of Queensland {.unlisted .unnumbered}  
#### [Hanny Rivera, PhD](https://www2.whoi.edu/staff/hrivera/), Woods Hole Oceanographic Institute  {.unlisted .unnumbered}
#### Jecy Klinkam, College of William and Mary (Intern) {.unlisted .unnumbered}
#### [Joanna Griffiths, PhD](https://joannasgriffiths.wordpress.com/), UC-Davis {.unlisted .unnumbered}
#### [Joe McGirr, PhD](https://joemcgirr.github.io/), UC-Davis {.unlisted .unnumbered}
#### Kaitlin Macaranas, CSU-Bakersfield (Intern) {.unlisted .unnumbered}
#### Megan Guidry, University of Rhode Island {.unlisted .unnumbered}
#### Natalie Ameral, University of Rhode Island & RI Dept. of Environmental Management {.unlisted .unnumbered}  
#### Nicolas Lou, Cornell University {.unlisted .unnumbered}
#### [Sara Schaal](https://schaals.wixsite.com/smschaal), Northeastern University {.unlisted .unnumbered} 
#### [Sara Wuitchik, PhD](https://sjswuitchik.weebly.com/), Harvard University {.unlisted .unnumbered}

>>>>

* [Multifactorial RNAseq analyses](DGE_comparison_v2.html)  
* Align and count RNAseq reads (coming soon!)
    * De novo, reference-based, and Tag-seq methods

@@

#### [Katherine Silliman, PhD](https://ksilliman.weebly.com/), South Carolina Dept. of Natural Resources - sillimank[at]dnr.sc.gov {.unlisted .unnumbered}
#### [Sam Bogan](http://www.hofmannlab.com/sam-bogan.html), UC-Santa Barbara -snbogan[at]ucsb.edu {.unlisted .unnumbered}
#### [Alan Downey-Wall](https://cos.northeastern.edu/people/alan-downey-wall/), Northeastern University {.unlisted .unnumbered}
#### Chris Mantegna, University of Washington (Intern)  {.unlisted .unnumbered}
#### [Danielle Davenport](https://www.researchgate.net/profile/Danielle-Davenport-2), University of Queensland {.unlisted .unnumbered}  
#### [Hanny Rivera, PhD](https://www2.whoi.edu/staff/hrivera/), Woods Hole Oceanographic Institute  {.unlisted .unnumbered}
#### Jecy Klinkam, College of William and Mary (Intern) {.unlisted .unnumbered}
#### [Joanna Griffiths, PhD](https://joannasgriffiths.wordpress.com/), UC-Davis {.unlisted .unnumbered}
#### [Joe McGirr, PhD](https://joemcgirr.github.io/), UC-Davis {.unlisted .unnumbered}
#### Kaitlin Macaranas, CSU-Bakersfield (Intern) {.unlisted .unnumbered}
#### Megan Guidry, University of Rhode Island {.unlisted .unnumbered}
#### Natalie Ameral, University of Rhode Island & RI Dept. of Environmental Management {.unlisted .unnumbered}  
#### Nicolas Lou, Cornell University {.unlisted .unnumbered}
#### [Sara Schaal](https://schaals.wixsite.com/smschaal), Northeastern University {.unlisted .unnumbered} 
#### [Sara Wuitchik, PhD](https://sjswuitchik.weebly.com/), Harvard University {.unlisted .unnumbered}
